The MACE is essentially unidimensional: Bifactor models of the `Maltreatment and Abuse Chronology of Exposure' scale

### 2022-07-28 manuscript/main.tex
manuscript updates (pt 2, today)

\title{The `Maltreatment and Abuse Chronology of Exposure' (MACE) scale is essentially unidimensional}
\shorttitle{Bifactor models of MACE}
\author{Samuel Zorowitz$^1$, Lauri Tuominen$^{2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$The Royal’s Institute of Mental Health Research, University of Ottawa, Canada}

@@

To evaluate the reliability of the MACE total and subscale scores, we calculated on several model-based reliability indices using the standardized factor loadings from the bifactor models. First, we calculated coefficient omega ($\omega$; \citealt{mcdonald1999test}), which quantifies the proportion of reliable score variance attributable to all factors (i.e. group and specific factors). Next we calculated the coefficient omega hierarchical ($\omega_h$) and coefficient omega subscale ($\omega_s$; \citealt{reise2013applying, rodriguez2016evaluating}). The $\omega_h$ index quantifies the fraction of variance in total scores that can be attributed to the individual differences on the general factor. By contrast, the $\omega_s$ index quantifies the fraction of variance in a subscale score that can be attributed to individual differences on a specific factor after controlling for the general factor. The $\omega_h$/$\omega_s$ values facilitate interpretation of the total and subscale scores. Values of $\omega_h$ approaching 1 indicate the general factor is the most dominant source of reliable variance in a total score. When $\omega_h$ is greater than 0.80, the total score on the scale may be considered essentially unidimensional \citep{rodriguez2016applying}. Similarly, values of $\omega_s$ approaching 1 increasingly indicate that individual differences on a subscale score reflect a specific factor and not the general factor. \cite{canivez2016bifactor} suggests that an acceptable $\omega_s$ value is 0.50 and that >0.70 is desirable. In contrast, $\omega_s$ values substantially smaller than 0.50 prevent meaningful interpretation of a subscale as reflecting a specific factor \citep{gignac2013bifactor}.

In order to assess the unidimensionality of the MACE scale, we also calculated for each bifactor model and sample the explained common variance (ECV; \citealt{sijtsma2009use}) and the percent uncontaminated correlations (PUC; \citealt{reise2013multidimensionality}). The ECV is the ratio of the variance explained by the general factor divided by the total common variance of a scale, and thus quantifies the contribution of the general factor relative to the specific factors. As ECV values approach 1, general factor loadings are increasingly expected to resemble those that would be obtained by a one-dimensional model; ECV values greater than 0.70 are often indicative of essential unidimensionality \citep{rodriguez2016applying}. The PUC --- which quantifies how many inter-item correlations are accounted for only by the general factor --- is an important moderator of the ECV. When the PUC increases, the ECV is less important for evaluating the risk of bias when fitting a unidimensional model to data with a latent bifactor structure; when PUC is greater than 0.80, there is a low risk of bias when a multidimensional scale is treated as unidimensional \citep{reise2013multidimensionality}. As an additional check, we calculated for each bifactor model the relative parameter bias, or the difference between an item's loading in the unidimensional  model and its corresponding general factor loading the bifactor, divided by the general factor loading in the bifactor. Parameter bias less than 10–15\% is acceptable \citep{muthen1987structural}.

Though of secondary interest, we also calculated the H index to assess the construct construct replicability of each factor \citep{hancock2001rethinking}. The H index is an estimate of how well a set of items represents a latent variable. Larger H values (e.g. $> 0.70$) indicate a well-defined latent variable, with factor loadings more likely to be stable across studies; conversely, low H values suggest a poorly defined latent variable, where the pattern of factor loadings are liable to change across studies \citep{hancock2001rethinking}.

@@

In addition to confirmatory factor analysis, we also performed exploratory factor analysis (EFA) on the endorsement data from both samples to see what factor structure would emerge with no \emph{a priori} restrictions \citep{schmitt2018selecting}. We elected to extract two- and three-factor solutions based to match the number of factors in the two bifactor S-1 models. That is, we wanted to see if a two-factor solution would resemble threat and deprivation factors; and if a three-factor solution would resemble family maltreatment, peer victimization, and reverse-scoring factors.

EFA was performed using the \textit{lavaan} (v0.6.11; \citealt{lavaan}) package available in R. Factors were extracted using the geomin \citep{yates1987multivariate} and cf-quartimax rotation criteria \citep{crawford1970general}, which were selected in order to extract simple factor structures with few cross-loadings. We used two rotation criteria in order to examine the generalizability of the factor solutions across oblique and orthogonal rotations.

@@

The goodness-of-fit indices for each model and sample are summarized in Table \ref{table:diagnostics}. According to the traditional indices, each model provided at least an acceptable fit to the response data. All RMSEA value were <0.06 and the majority of CFI and TLI values were >0.95. The only exception were for the fits of models 1 and 3 to the original sample data, which exhibited only acceptable fits (i.e. CFI/TLI > 0.90). In general, model fits were better for the replication sample data than for the original sample. Furthermore, all of the posterior predictive p-values corresponding to the $\chi^2_{NC}$ discrepancy measure were within the acceptable range. Thus all models were sufficiently able to reproduce the distribution of overall severity scores for each sample.

>>>>

\title{The MACE is essentially unidimensional: Bifactor models of the `Maltreatment and Abuse Chronology of Exposure' scale}
\shorttitle{Bifactor models of MACE}
\author{Samuel Zorowitz$^1$, Lauri Tuominen$^{2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$The Royal’s Institute of Mental Health Research, University of Ottawa, Canada}

@@

To evaluate the reliability of the MACE total and subscale/composite scores, we computed several model-based reliability indices using the factor loading estimates from the bifactor models. First, we calculated coefficient omega ($\omega$; \citealt{mcdonald1999test}), which quantifies the proportion of reliable score variance attributable to all factors (i.e. both group and specific factors). Next we calculated coefficient omega hierarchical ($\omega_h$) and coefficient omega subscale ($\omega_s$;  \citealt{reise2013applying, rodriguez2016evaluating}). Whereas $\omega_h$ measures the fraction of total score variance attributable to the general factor, $\omega_s$ measures the fraction of subscale score variance attributable to a specific factor after controlling for the general factor. The values of $\omega_h$/$\omega_s$ facilitate interpretation of total and subscale scores. Values of $\omega_h$ approaching 1 indicate the general factor is the primary source of reliable variance in a total score. When $\omega_h$ > 0.80, the total score may be considered an essentially unidimensional reflection of the general factor (even if the data are multidimensional;  \citealt{rodriguez2016applying}). So too values of $\omega_s$ approaching 1 indicate a specific factor, and not the general factor, is the primary source of reliable variance in a subscale score. \cite{canivez2016bifactor} suggests that an acceptable $\omega_s$ value is 0.50 and that >0.70 is desirable. In contrast, values of $\omega_s$ smaller than 0.50 preclude meaningful interpretation of subscale scores as reflecting a specific factor \citep{gignac2013bifactor}.

To assess the unidimensionality of the MACE scale, we also calculated the explained common variance (ECV; \citealt{sijtsma2009use}) and the proportion of uncontaminated correlations (PUC; \citealt{reise2013multidimensionality}). The ECV is the ratio of the variance explained by the general factor divided by the total common variance of a scale; it therefore measures the contribution of the general factor relative to the specific factors. As ECV values approach 1, the general factor loadings of a bifactor model are expected to resemble those that would be obtained by a one-dimensional model. ECV values greater than 0.70 often indicate a scale is essential unidimensional \citep{rodriguez2016applying}. In turn, the PUC quantifies how many inter-item correlations are accounted for only by the general factor and is an important moderator of the ECV. As the PUC increases, ECV is less important for evaluating the risk of bias when fitting a unidimensional model to data with a latent bifactor structure. When PUC > 0.80, there is a low risk of bias when a multidimensional scale is treated as unidimensional \citep{reise2013multidimensionality}. As an additional check, we calculated the relative parameter bias (RPB), which is the difference between an item's loading in a unidimensional model and its corresponding general factor loading the bifactor, divided by the general factor loading in the bifactor. Parameter bias less than 10–15\% is acceptable \citep{muthen1987structural}.

Finally, we calculated the H index to assess the construct replicability of each factor \citep{hancock2001rethinking}. The H index is an estimate of how well a set of items represents a latent variable. H values closer to 1 indicate a well-defined latent variable with factor loadings more likely to be stable across studies. In contrast, small H values suggest a poorly defined latent variable where factor loadings are liable to change across studies. Because we are primarily focused on measuring the reliability of MACE total and subscale/composite scores, we report the H values below but do not discuss them further.

@@

In addition to confirmatory factor analysis, we also we also performed exploratory factor analysis (EFA) on the response data from each sample separately. The goal of the EFA was to observe what factor structure would emerge with no \emph{a priori} restrictions \citep{schmitt2018selecting}. We decided to extract two- and three- factor solutions based on the total number of factors in the bifactor S-1 models. We were particularly interested in seeing if a two-factor solution would produce the factors from model 3 (i.e. threat and deprivation); and if a three-factor solution would produce the three factors from model 4 (i.e. maltreatment by adults, peer victimization, reverse-scoring).

EFA was performed using the \textit{lavaan} (v0.6.11; \citealt{lavaan}) package available in R. Factors were extracted using the geomin \citep{yates1987multivariate} and cf-quartimax rotation criteria \citep{crawford1970general}, which were selected in order to extract simple factor structures with few cross-loadings. We used two rotation criteria in order to examine the generalizability of the factor solutions across multiple rotations.

@@

The goodness-of-fit indices for each model and sample are summarized in Table \ref{table:diagnostics}. According to the traditional indices, each model provided at least an acceptable fit to the response data. All RMSEA value were <0.05 and the majority of CFI and TLI values were >0.95. The only exception were for the fits of models 1 and 3 to the original sample data, which exhibited only acceptable fits (i.e. CFI/TLI > 0.90). In general, model fits were better for the replication sample data than for the original sample. Furthermore, all of the posterior predictive p-values corresponding to the $\chi^2_{NC}$ discrepancy measure were within the acceptable range. Thus all models were sufficiently able to reproduce the distribution of overall severity scores for each sample.
